# General Statistics
  
In this course, we will require the ecommerce data set that can be downloaded from [here](https://www.kaggle.com/datasets/shrishtimanja/ecommerce-dataset-for-data-analysis/data). 

Lets load the data set
```{r}
ecommerce <- read.csv("data/ecommerce.csv")

# First few rows of the data set 
head(ecommerce)
```

We will explore the fundamentals of statistics using the data set that is packed with real-life insights on customers and their purchasing behavior. At the end of this course unit you will be able to understand;

- Measure of spread and central tendency
  - Mean(the average)
  - Mode
  - Quartiles
  - Range(IQR, maximum and minimum)

- Data Visualizations 
  - Relationship(scatter plot and heatmaps)
  - Trend(line chart)
  - Distribution(histograms and density plots)
  - Comparison(bar charts)
  - Composition (pie charts)
  - Outliers (boxplots and violin plots)
  
- Inferential Statistics
  - Hypothesis testing 
  - T-tests( one sample, anova, chi-square)

## Measure of Spread and Central Tendency

a. **Arithmetic mean**

It is also referred to as the average. It is the sum of all values divided by the number of values in the set. R has a function `mean` to calculate the average. 

**Formula for mean**

Lets calculate the average net amount spent on the ecommerce store. 
```{r}
# Calculate the mean net amount
mean(ecommerce$Net.Amount)
```

The average amount spent on the e-commerce store by all the customers was 2875.95 INR. 

b. **Mode**

The mode shows the most frequent value, thereby pinpoint the most popular categories in a data set. In the commerce data set we will find the most popular category and age group in the data set. 

Find the most popular category
```{r}
# Calculate the mode for 'ProductCategory'
table(ecommerce$Product.Category)
which.max(table(ecommerce$Product.Category))
```

c. **Quartiles**

**____Come back later_____**









