# Simple Linear Regression

Welcome to the world of Simple Linear Regression! 🎉 This statistical technique is super handy when you want to explore the relationship between two continuous variables. Essentially, it helps us predict the value of one variable based on the value of another.

For example, imagine you want to predict a student's exam score based on the number of hours they studied. Here, the hours studied are the **independent variable** (or predictor), and the exam score is the **dependent variable** (or response).

**What is Simple Linear Regression?**

In simple linear regression, we fit a straight line (called the regression line) through the data points. This line is defined by the equation:

$$y = mx + b$$

Where:

- $y$ is the predicted value (dependent variable).
- $m$ is the slope of the line (how much $y$ changes for a unit change in $x$).
- $x$ is the independent variable.
- $b$ is the y-intercept (the value of $y$ when $x$ is 0).

Let's use the built-in `mtcars` data set in R to demonstrate how to perform simple linear regression.

Load the data set
```{r}
# Load the mtcars dataset
data(mtcars)
# View the first few rows of the dataset
head(mtcars)
```

Fit the simple linear regression model that will predict `mpg` (miles per gallon) based on `wt` (the weight of the car).
```{r}
# Fit the linear regression model
model <- lm(mpg ~ wt, data = mtcars)
```

- Get the model summary to get important information about the model we just fitted. 
```{r}
# Get the summary of the model
summary(model)
```

- Plot the data and regression line to understand the relationship
```{r}
# Plot the data points
plot(mtcars$wt, mtcars$mpg, main = "Simple Linear Regression", 
     xlab = "Weight of the Car (wt)", ylab = "Miles Per Gallon (mpg)", 
     pch = 19, col = "blue")

# Add the regression line
abline(model, col = "red")
```

Predictions can be made based on the data. Lets predict the mpg for car that weighs 3.5 tons
```{r}
# Predict mpg for a car that weighs 3.5 tons
new_data <- data.frame(wt = 3.5)
predicted_mpg <- predict(model, new_data)
print(paste("Predicted MPG for a car weighing 3.5 tons:", round(predicted_mpg, 2)))
```

## Basics of Wilkinson-Rogers Notation (y ~ x), Linear Regression


## Scatterplots with Regression Lines, Reading lm() Output

## Confidence Intervals for Regression Coefficients, Testing Coefficients


## Identifying Points in a Plot