# Simple Linear Regression

Welcome to the world of Simple Linear Regression! 🎉 This statistical technique is super handy when you want to explore the relationship between two continuous variables. Essentially, it helps us predict the value of one variable based on the value of another.

For example, imagine you want to predict a student's exam score based on the number of hours they studied. Here, the hours studied are the **independent variable** (or predictor), and the exam score is the **dependent variable** (or response).

**What is Simple Linear Regression?**

Simple linear regression is essentially about finding the line that best describes the relationship between two variables(independent and dependent variables) in your data. The line is called the **regression line**, and it helps us make predictions. 

## Basics of Wilkinson-Rogers Notation (y ~ x), Linear Regression

In simple linear regression, we fit a straight line (called the regression line) through the data points. This line is defined by the equation:

$$y = mx + b$$

Where:

- $y$ is the predicted value (dependent variable).
- $m$ is the slope of the line (how much $y$ changes for a unit change in $x$).
- $x$ is the independent variable.
- $b$ is the y-intercept (the value of $y$ when $x$ is 0).

<span style="color: orange;">**Try it!**</span> 

Lets say, we have a data set of age and height of youngsters as below here:
```
age = c(5, 10, 15, 20, 25)
height = c(110, 130, 150, 160, 170)
```

We can use the age to predict the height where age is the independent variable and height is the dependent variable(depends on age). The equation is written like this in R;
```
height ~ age
```

in short `independent_variable ~ dependent_variable`. Now, lets work it out using the `lm()` function
```{r}
# Creating a sample data set
age = c(5, 10, 15, 20, 25)
height = c(110, 130, 150, 160, 170)

# Fitting a linear regression data set
model_1 <- lm(height ~ age)
```

The `lm()` function stands fir 'linear model` and it finds the best line to describle the relationship between height and age. 
```{r}
summary(model_1)
```

<span style="color: orange;">**Real world challenge**</span> 

Let's use the built-in `mtcars` data set in R to demonstrate how to perform simple linear regression.

Load the data set
```{r}
# Load the mtcars dataset
data(mtcars)
# View the first few rows of the dataset
head(mtcars)
```

Fit the simple linear regression model that will predict `mpg` (miles per gallon) based on `wt` (the weight of the car).
```{r}
# Fit the linear regression model
model_2 <- lm(mpg ~ wt, data = mtcars)
```

- Get the model summary to get important information about the model we just fitted. 
```{r}
# Get the summary of the model
summary(model_2)
```

## Scatterplots with Regression Lines, Reading lm() Output

Now that we have a model in place lets plot the data and regression line to understand the relationship. You remember we have just worked on a case of where we use the age of youngsters to find their height? That's fine! Lets plot a scatter plot and regression line to visualize the relationship between age and height.
```{r}
# Scatter plot of age vs height 
plot(age, height,
     main = "Height vs Age", 
     xlab = "Age",
     ylab = "Height",pch=19,
     col = "blue"
       )
```

The scatter plot shows that the youngsters tend to be taller as they get older. Now lets add a regression line to the data to see how well it fits the data;
```{r}
# Scatter plot of age vs height 
plot(age, height,
     main = "Height vs Age", 
     xlab = "Age",
     ylab = "Height",pch=19,
     col = "blue"
       )

# Adding a regression line 
abline(model_1, col="red", lwd=2)
```

The regression line helps us visually understand the trend. In our case here, the regression line closely follows the data points, therefore, our model is a good fit!

<span style="color: orange;">**Real World example**</span> 

We will still consider the model that we created above for the `mtcars` data set. Lets plot a scatter plot and fit a regression line based on the model that we have just created. 

```{r}
# Plot the data points
plot(mtcars$wt, mtcars$mpg, 
     main = "Simple Linear Regression", 
     xlab = "Weight of the Car (wt)", 
     ylab = "Miles Per Gallon (mpg)", 
     pch = 19, col = "blue")

# Add the regression line
abline(model_2, col = "red")
```

Predictions can be made based on the data. Lets predict the mpg for car that weighs 3.5 tons
```{r}
# Predict mpg for a car that weighs 3.5 tons
new_data <- data.frame(wt = 3.5)
predicted_mpg <- predict(model_2, new_data)
print(paste("Predicted MPG for a car weighing 3.5 tons:", round(predicted_mpg, 2)))
```

## Confidence Intervals for Regression Coefficients, Testing Coefficients


## Identifying Points in a Plot