--- 
title: "Statistical Computing"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
---

# Reviews
## Data Types and Structures 
### Data Types 
There are different kinds of values in R that can be manipulated in variables. They most commonly used are; strings, numerics(integers and floats) and boolean values. The function `class` can be used to find the data type.

<span style="color: orange;">**Try it!**</span> 

Before diving deep into data types, lets create a value with a random value for instance `age` and finds its data type. 
```{r}
age <- 27
class(age)
```

The `age` is a `"numeric"` data type variable, interesting?

Lets explore different data types and their examples; 

1. **Integers: ** These are whole numbers without dev=cimal point(e.g., 10, -5). In R, it is specified with the `L` suffix like `10L`. 
2. **Floats: ** These are numbers with decimal points(e.g. 3.14, -2.718). R refers them as numerics. 
3. **Boolean(Logical): ** True or False values, represented as `TRUE` or `FALSE` in R. They are crucial in conditional statements. 
4. **Strings(Character): ** These are text values enclosed in quotes(e.g. `"Hello world"` , names like `"John", "Mustafa", "Patel"`, variable names like `"age", "gender", "salary"`)

You will often deal with mixed data types when analyzing real-world data sets therefore understanding these will help you handles any data set!

<span style="color: orange;">**Examples**</span>

Lets have some fun! We will create different variables and find their data types;

i. `age <- 34L`
```{r}
age <- 34L
class(age)
```
ii. `weight <- 68.2`
```{r}
weight <- 68.2
class(weight)
```
iii. `name <- "Mustafa"`
```{r}
name <- "Mustafa"
class(name)
```
iv. `is_winter <- FALSE`
```{R}
is_winter <- FALSE
class(is_winter)
```

You see how simple it is to find the data type of different variables in R! Remember the `class` function returns any number whether with decimal or whole as `"numeric"`. It only returns `"integer"` when there is a suffix `L`. 

<span style="color: green;">**Practical Exercise**</span>

Try out the practical exercise below to test your understanding in data types 

i. Find the data type of `98.03` using `class()` function.
ii. Assign the value `98.03` to variable `height` and find data type of height.
iii. There are 27 goats in a field, assign the quantity of goats to a variable `goats` and find the data type of the variable `goats`. Remember to add suffix `L` to  the value `27`.
iv. Find the data type of the value `"school"` using the `class()` function.
v. Assign your first name to a variable `firstname` and find its data type. **Remember to enclose it in quotation marks**
vi. Create a variable `is_student` and assign it the value `TRUE`. Use the `class()` function to find its data type. 

_______________________________________________________________________
<span style="color: brown;">**Solution**</span> 

i. Find the data type of `98.03` using `class()` function.
```{r}
class(98.03)
```
ii. Assign the value `98.03` to variable `height` and find data type of height.
```{r}
height <- 98.03
class(height)
```
iii. There are 27 goats in a field, assign the quantity of goats to a variable `goats` and find the data type of the variable `goats`. Remember to add suffix `L` to  the value `27`.
```{r}
goats <- 27L
class(goats)
```
iv. Find the data type of the value `"school"` using the `class()` function.
```{r}
class("school")
```
v. Assign your first name to a variable `firstname` and find its data type. **Remember to enclose it in quotation marks**
```{r}
firstname <- "Bryant" # Any name will work
class(firstname)
```
vi. Create a variable `is_student` and assign it the value `TRUE`. Use the `class()` function to find its data type. 
```{r}
is_student <- TRUE
class(is_student)
```
<span style="color: brown;">**________________________________________________________________________________**</span>

### Data Structures 
This is the organization of data into or multiple data values in specific structures, they include vectors, matrix and data frames. Lets explore the mentioned data structures and their examples;

1. **Vector: ** This is a sequence of elements of the same data types(e.g., ``c(1, 2, 3)` is a numeric vector)
2. **Matrix: ** This is a two-dimensional data structure with rows and columns, where all elements are of the same type(e.g. numbers).
3. **Data Frames: ** This is the most common R data structure for handling tabular data(like an excel sheet). A data frame can contain different data types in each column unlike matrices and vectors. 

Data frames are central to real-world data analysis. You will work with them to analyze, transform, and visualize data sets, whether you are calculating averages or identifying trends. The `is.vector`, `is.matrix` and `is.data.frame` functions are used to confirm if the variable in question is a vector, matrix or data frame respectively. 

<span style="color: orange;">**Examples**</span>

Lets have some fun! We will create different data structures and find their data types: 

- Create a vector, `marks` to store the values, `23, 67, 98, 34, 98, 21`. Print the vector to the console and use `is.vector` function to confirm if its a actually a vector. 
```{r}
marks = c(23, 67, 98, 34, 98, 21)
print(marks) # print to the console
is.vector(marks) # find its data structure
```
- Create a matrix with values from 1 to 9 and use the `is.matrix` function to find to confirm if its really a matrix.
```{r}
vector1 = seq(1, 9) 

# Convert to matrix
## create by column 
m1=matrix(vector1, ncol=3)
print(m1) # print the matrix to the console 

is.matrix(m1) # confirms if its really a matrix 
```
- Create a `data.frame` from the above matrix. Add the column names as `"A", "B", "C"`. Confirm if its really a matrix. 
```{r}
var_names <- c("A", "B", "C") # vector to store variable names 

m1_df <- data.frame(var_names, m1) # create the data frame
print(m1_df) #print to the console 

is.data.frame(m1_df) #confirms if its really a data.frame. 
```

<span style="color: green;">**Practical Exercise**</span> 

## Data Importing and Exporting 
- Data structures and how to import data into R

## Data Cleaning 


## Data Visualization
- A little bit of data visualization
- 