--- 
title: "Statistical Computing"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
---

# Reviews
## Data Types and Structures 
### Data Types 
There are different kinds of values in R that can be manipulated in variables. They most commonly used are; strings, numerics(integers and floats) and boolean values. The function `class` can be used to find the data type.

<span style="color: orange;">**Try it!**</span> 

Before diving deep into data types, lets create a value with a random value for instance `age` and finds its data type. 
```{r}
age <- 27
class(age)
```

The `age` is a `"numeric"` data type variable, interesting?

Lets explore different data types and their examples; 

1. **Integers: ** These are whole numbers without dev=cimal point(e.g., 10, -5). In R, it is specified with the `L` suffix like `10L`. 
2. **Floats: ** These are numbers with decimal points(e.g. 3.14, -2.718). R refers them as numerics. 
3. **Boolean(Logical): ** True or False values, represented as `TRUE` or `FALSE` in R. They are crucial in conditional statements. 
4. **Strings(Character): ** These are text values enclosed in quotes(e.g. `"Hello world"` , names like `"John", "Mustafa", "Patel"`, variable names like `"age", "gender", "salary"`)

You will often deal with mixed data types when analyzing real-world data sets therefore understanding these will help you handles any data set!

<span style="color: orange;">**Examples**</span>

Lets have some fun! We will create different variables and find their data types;

i. `age <- 34L`
```{r}
age <- 34L
class(age)
```
ii. `weight <- 68.2`
```{r}
weight <- 68.2
class(weight)
```
iii. `name <- "Mustafa"`
```{r}
name <- "Mustafa"
class(name)
```
iv. `is_winter <- FALSE`
```{R}
is_winter <- FALSE
class(is_winter)
```

You see how simple it is to find the data type of different variables in R! Remember the `class` function returns any number whether with decimal or whole as `"numeric"`. It only returns `"integer"` when there is a suffix `L`. 

<span style="color: green;">**Practical Exercise**</span>

Try out the practical exercise below to test your understanding in data types 

i. Find the data type of `98.03` using `class()` function.
ii. Assign the value `98.03` to variable `height` and find data type of height.
iii. There are 27 goats in a field, assign the quantity of goats to a variable `goats` and find the data type of the variable `goats`. Remember to add suffix `L` to  the value `27`.
iv. Find the data type of the value `"school"` using the `class()` function.
v. Assign your first name to a variable `firstname` and find its data type. **Remember to enclose it in quotation marks**
vi. Create a variable `is_student` and assign it the value `TRUE`. Use the `class()` function to find its data type. 

_______________________________________________________________________
<span style="color: brown;">**Solution**</span> 

i. Find the data type of `98.03` using `class()` function.
```{r}
class(98.03)
```
ii. Assign the value `98.03` to variable `height` and find data type of height.
```{r}
height <- 98.03
class(height)
```
iii. There are 27 goats in a field, assign the quantity of goats to a variable `goats` and find the data type of the variable `goats`. Remember to add suffix `L` to  the value `27`.
```{r}
goats <- 27L
class(goats)
```
iv. Find the data type of the value `"school"` using the `class()` function.
```{r}
class("school")
```
v. Assign your first name to a variable `firstname` and find its data type. **Remember to enclose it in quotation marks**
```{r}
firstname <- "Bryant" # Any name will work
class(firstname)
```
vi. Create a variable `is_student` and assign it the value `TRUE`. Use the `class()` function to find its data type. 
```{r}
is_student <- TRUE
class(is_student)
```
<span style="color: brown;">**________________________________________________________________________________**</span>

### Data Structures 
This is the organization of data into or multiple data values in specific structures, they include vectors, matrix and data frames. Lets explore the mentioned data structures and their examples;

1. **Vector: ** This is a sequence of elements of the same data types(e.g., ``c(1, 2, 3)` is a numeric vector)
2. **Matrix: ** This is a two-dimensional data structure with rows and columns, where all elements are of the same type(e.g. numbers).
3. **Data Frames: ** This is the most common R data structure for handling tabular data(like an excel sheet). A data frame can contain different data types in each column unlike matrices and vectors. 

Data frames are central to real-world data analysis. You will work with them to analyze, transform, and visualize data sets, whether you are calculating averages or identifying trends. The `is.vector`, `is.matrix` and `is.data.frame` functions are used to confirm if the variable in question is a vector, matrix or data frame respectively. 

<span style="color: orange;">**Examples**</span>

Lets have some fun! We will create different data structures and find their data types: 

- Create a vector, `marks` to store the values, `23, 67, 98, 34, 98, 21`. Print the vector to the console and use `is.vector` function to confirm if its a actually a vector. 
```{r}
marks = c(23, 67, 98, 34, 98, 21)
print(marks) # print to the console
is.vector(marks) # find its data structure
```
- Create a matrix with values from 1 to 9 and use the `is.matrix` function to find to confirm if its really a matrix.
```{r}
vector1 = seq(1, 9) 

# Convert to matrix
## create by column 
m1=matrix(vector1, ncol=3)
print(m1) # print the matrix to the console 

is.matrix(m1) # confirms if its really a matrix 
```
- Create a `data.frame` from the above matrix. Add the column names as `"A", "B", "C"`. Confirm if its really a matrix. 
```{r}
var_names <- c("A", "B", "C") # vector to store variable names 

m1_df <- data.frame(var_names, m1) # create the data frame
print(m1_df) #print to the console 

is.data.frame(m1_df) #confirms if its really a data.frame. 
```

<span style="color: green;">**Practical Exercise**</span> 

Try out the exercise below to test your understanding in R data structures;

i. Create a vector named `height` with the values `120.1, 118, 123.4, 130.8, 115.2`. Use the `is.vector` to confirm that the created variable is a vector.
ii. Use `length()` function to count the number of elements in the vector. 
iii. Create a matrix `m1` from the vector `v1` where `v1 <- seq(1, 12)` with three columns. Use the `is.matrix` function to confirm if the said variable is a matrix. 
iv. Access the third column by running the command `m1[, 3]`.
v. Access the second row by running the command `m1[2,]`. 
vi. Create a data frame `students_df` with the columns `"Name"`, `"Age"`, and `"Marks"` for three students. Where `Name <- c("Pragya", "Thomas", "Ali")`, `Age <- c(21, 19, 23)` and `Marks <- c(68, 72, 67)`. 

_______________________________________________________________________
<span style="color: brown;">**Solution**</span> 

i. Create a vector named `height` with the values `120.1, 118, 123.4, 130.8, 115.2`. Use the `is.vector` to confirm that the created variable is a vector.
```{r}
height <- c(120.1, 118, 123.4, 130.8, 115.2)
is.vector(height)
```

ii. Use `length()` function to count the number of elements in the vector.
```{r}
length(height) # count the number of elements
```

iii. Create a matrix `m1` from the vector `v1` where `v1 <- seq(1, 12)` with three columns. Use the `is.matrix` function to confirm if the said variable is a matrix. 
```{r}
v1 <- seq(1, 12) # Create vector v1

m1 <- matrix(v1, ncol=3) # create a matrix from the vector
m1 

is.matrix(m1) # Confirm if its a matrix
```

iv. Access the third column by running the command `m1[, 3]`.
```{r}
m1[, 3] # Access the third columns
```

v. Access the second row by running the command `m1[2,]`. 
```{r}
m1[2,] # Access of the second row
```

vi. Create a data frame `students_df` with the columns `"Name"`, `"Age"`, and `"Marks"` for three students. Where `Name <- c("Pragya", "Thomas", "Ali")`, `Age <- c(21, 19, 23)` and `Marks <- c(68, 72, 67)`. 
```{r}
student_df <- data.frame(
  "Name"= c("Pragya", "Thomas", "Ali"),
  "Age"=c(21, 19, 23),
  "Marks"=c(68, 72, 67)
)

student_df
```

<span style="color: brown;">**________________________________________________________________________________**</span>

## Data Importing and Exporting 
Importing and exporting data is the foundation of data analysis workflows. The main two types of data files used are CSV and excel files. 

- **CSV Files: ** R can easily import CSV files using `read.csv("filename.csv")`. The CSV is one of the most common formats you will encounter.
- **Excel Files: ** For excel files, you can use the `readxl` package with the function `read_excel`. 

<span style="color: orange;">**Try it: **</span>

Let's have some fun by importing;

i. From CSV file 
```{r}
m1_imported <- read.csv("data/m1.csv") # import the csv data set
m1_imported # Display the data
```
ii. From Excel file
```{r}
library(readxl)

students_imported <- read_excel("data/students.xlsx") # Import the data
students_imported # Display the data
```

After data wrangling, manipulation and processing, the end product(processed data) can be saved for further use. The data can also be shared to others. Lets explore how export the CSV and Excel files 

- **To CSV: ** You can save your data to CSV format using `write.csv(data, "filepath.csv")`.
- **To Excel** To write to Excel, you can use `write.xlsx(data, "filepath.xlsx")` from the `openxlsx` package. 

<span style="color: orange;">**Try it: **</span>

Lets export the previously imported data set locally 

i. To CSV
```{r}
write.csv(m1_imported, "data/m1_exported.csv") # Write the data set locally
```
ii. Excel File 
```{r}
library(openxlsx)

write.xlsx(students_imported, "data/students_exported.xlsx")
```

<span style="color: green;">**Practical Exercise**</span>

## Data Cleaning 

Before you analyze data, it is crucial to ensure that it is clean. Here are some common issues in data cleaning;

- **Null Values: ** Missing data can distort your analysis. Functiions like `is.na()` and `na.omit()` are used to detect and remove null values respectively. 
- **Duplicated Records: ** Duplicates can cause bias in results and they can detected using the `duplicated()` function. This duplicated records can be removed by `unique()` function from R or `distinct()` from `dplyr` package. 
- **Outliers: ** These are extreme values that don't follow the general trend. The use of summary statistics(specifically IQR) and boxplots can be used to cap these values based on the context. 


Data cleaning is like polishing a diamond-it ensures the data is ready for analysis, free from distortions like missing values or outliers that can skew your insights. 

<span style="color: orange;">**Try it: **</span> 

Lets have some fan! We will create a random data set, identify all data issues and address them by cleaning. 

- Create the data set 
```{r}
#Remember to create the data set
```

<span style="color: green;">**Practical Exercise**</span>

## Data Visualization

**____What is Data Visualization____** `ggplot` is one of the most popular and flexble data visualization libraries in R. It follows the **grammar of graphics** philosophy, allowing you o build plots in layers. Here are some of the basic plots in data visualization;

- **Scatter Plots: ** used to visualize the relationship between two variables in R.
- **Bar Charts: ** used to compare categorical data. 
- **Histograms: ** used to represent distribution of a single continuous variable. 

Visualizations are powerful tools that help you see patterns and insights that raw data might hide. A well-made plot can communicate your findings more effectively than numbers alone. 

<span style="color: orange;">**Try it: **</span>

Lets use the above data set that we cleaned to plot simple charts in R using `ggplot` library. 

- Install the package if not installed
```
install.packages("ggplot2")
```

- Load the library
```{r}
library(ggplot2)
```

## <span style="color: green;">**Hands-on Exercises**</span> 

_______________________________________________________________________
<span style="color: brown;">**Solution**</span> 

<span style="color: brown;">**________________________________________________________________________________**</span>